---
title: "Predicting Expected Points Added in Football: A Linear Regression Approach"
title-block-banner: true
thanks: "Code and data supporting this analysis is available at: https://github.com/UofT-DailinLi/toronto-crime-report.git"
author: "Ben Li"
date: today
date-format: long
editor: visual
format: pdf
toc: TRUE
number-sections: true
bibliography: ref.bib
---

```{r echo=FALSE, message=FALSE}
# install.packages("caret")
# install.packages("nflverse")
library(nflverse)
# install.packages("nflreadr") 
library(nflreadr)
library(opendatatoronto)
library(tidyverse)
library(caret)
library(lmtest)
library(here)
library(knitr)
```

```{r echo=FALSE, message=FALSE}
qb_regular_season_stats <- load_player_stats(seasons = TRUE) |> filter(season_type == "REG" & position == "QB")

qb_data <- qb_regular_season_stats %>%
filter(week <= 9, season == 2023)
```

```{r echo=FALSE, message=FALSE}
# write_csv(
#  x = qb_data,
#  file = "qb_data_clean.csv")

```

```{r echo=FALSE, message=FALSE}
# Load necessary libraries


# Read the data
data = read_csv(
    file = here("outputs/data/qb_data_clean.csv"), show_col_types = FALSE)

# Selecting relevant columns
df <- data %>% select(passing_epa, attempts, completions, passing_yards, passing_tds, interceptions, sacks)

# Removing rows with NA values in any of the selected columns
df <- na.omit(df)

# Splitting the data into training and test sets
set.seed(123) # For reproducibility
splitIndex <- createDataPartition(df$passing_epa, p = .8, list = FALSE)
train <- df[splitIndex, ]
test <- df[-splitIndex, ]

# Building the linear regression model
model <- lm(passing_epa ~ attempts + completions + passing_yards + passing_tds + interceptions + sacks, data = train)

# Summary of the model to see coefficients and statistics


# Predicting on the test set
predictions <- predict(model, newdata = test)

# Evaluating the model with RMSE (Root Mean Squared Error)

rmse <- sqrt(mean((predictions - test$passing_epa)^2))


```

# Introduction

Football is an established industry in the United States with teams, fans, franchise merchandise, etc., and a complete economic system. Understanding the dynamics of a team can help a team succeed on the field. One of the important metrics when evaluating individual plays' effectiveness is Expected Points Added (EPA). This score effectively quantifies the impact of individual plays on scoring and can capture information beyond traditional methods. In this study, we predicted passing EPA through a linear regression model utilizing a variety of game-related variables. By focusing on quarterback performance, we analyzed the key factors that influence a team's ability to score.

# Data

Our dataset was obtained from Opentoronto and includes 15449 observations and 53 variables [@rOpenDataToronto]. We are only curious about the regional session QB position in the first nine weeks of 2023. Therefore, the number of observations dropped to 335. so the  Individual data from games inside the NFL are included, with a focus on quarterbacks (QB). The selected predictors include pass attempts, completions, passing yards, passing touchdowns, interceptions, and sacks. These variables were chosen because they are closely related to passing performance and may have a potential impact on EPA. The dataset was divided into a training set and a test set, randomly assigned in an 80-20 ratio, to develop the model and test its predictive ability. 

Data was cleaned and analyzed using the open source statistical programming language R [@r], and additional packages from `tidyverse` [@rTidyverse], `ggplot2` [@rGgplot2], ` nflverse ` [@nflverse], `here`[@rhere], `nflreadr`[@nflreadr], `caret`[@caret], `lmtest`[@lmtest] and `knitr` [@rKnitr]. The cleaned dataset contains 318 observations, and four variables will be shown below.


# Model

\begin{align*}
\text{passing\_epa} = \ & \beta_0 + \beta_1 \times \text{attempts} + \beta_2 \times \text{completions} + \\
& \beta_3 \times \text{passing\_yards} + \beta_4 \times \text{passing\_tds} + \beta_5 \times \text{interceptions} + \\
& \beta_6 \times \text{sacks} + \epsilon
\end{align*}

This is the linear model we used. In this equation $\beta_0$ = -0.79499 , $\beta_1$ = -0.82374, $\beta_2$ = 0.45150, $\beta_3$ = 0.09936, $\beta_4$ = 1.92008, $\beta_5$ = -3.37566 , $\beta_6$ = -1.91623 . 


```{r echo=FALSE, message=FALSE, fig.cap="Actual vs. Predicted passing EPA", fig.width=6, fig.height=4,label= "fig-act-pre",fig.pos="!h"}


library(ggplot2)

ggplot(test, aes(x = passing_epa, y = predictions)) +
  geom_point() +
  geom_line(aes(x = passing_epa, y = passing_epa), color = 'red') +
  theme_minimal() +
  labs(x = "Actual passing EPA", y = "Predicted passing EPA")

```

The scatterplot in @fig-act-pre  illustrates the relationship between the predicted and actual values of epa predicted by the regression model. Each point represents a value in a test dataset. The y-axis represents the predicted value from the model, the x-axis represents the actual value, and this red line represents the perfect prediction line, which means that the predicted and actual values are equal. The distance from these surrounding points to the line is the error. Overall, our model has relatively predictive solid power. However, the error may be likely due to the fact that the selected variables do not fully capture all the variations in the passing-EPA.


# Discussion
The resulting root mean square error is 4.2637 points, which indicates that, on average, the predicted values of our model do not differ much from the actual values of EPA. This error also suggests that the range of EPA should be analyzed using a case-by-case approach in soccer analysis. Moreover, key factors such as pass attempts and touchdowns are expected to be positively correlated with EPA, reflecting successful offensive plays. Conversely, interceptions and sacks may show a negative correlation.


\newpage
# Reference
